\textsc{TheHaloMod}: An online calculator for the halo model

### 2021-07-08 paper.tex
Update acknowledgment to have MK in it
Updates from Overleaf
Version for resubmission

\usepackage{array}
\usepackage{lineno}
\usepackage{xcolor}
\usepackage{lineno}
\usepackage[colorlinks=true,urlcolor=black,linkcolor=blue,citecolor=blue]{hyperref}
\ifthenelse{\boolean{useminted}}{\usepackage{minted}}{}
\usepackage{longtable,booktabs,tabu}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage{pbox}
\usepackage{amssymb}
\usepackage{layouts}
\usepackage{subfig}
\usepackage{microtype}
\usepackage{listings}

\usepackage{etoolbox}
\makeatletter
% Patch case where name and year have no delimiter
\patchcmd{\NAT@citex}
  {\@citea\NAT@hyper@{\NAT@nmfmt{\NAT@nm}\NAT@date}}
  {\@citea\NAT@nmfmt{\NAT@nm}\NAT@hyper@{\NAT@date}}
  {}% Do nothing if patch works
  {}% Do nothing if patch fails
% Patch case where name and year have basic delimiter
\patchcmd{\NAT@citex}
  {\@citea\NAT@hyper@{%
     \NAT@nmfmt{\NAT@nm}%
     \hyper@natlinkbreak{\NAT@aysep\NAT@spacechar}{\@citeb\@extra@b@citeb}%
     \NAT@date}}
  {\@citea\NAT@nmfmt{\NAT@nm}%
   \NAT@aysep\NAT@spacechar%
   \NAT@hyper@{\NAT@date}}
  {}% Do nothing if patch works
  {}% Do nothing if patch fails
% Patch case where name and year are separated by a prenote
\patchcmd{\NAT@citex}
  {\@citea\NAT@hyper@{%
     \NAT@nmfmt{\NAT@nm}%
     \hyper@natlinkbreak{\NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi}%
       {\@citeb\@extra@b@citeb}%
     \NAT@date}}
  {\@citea\NAT@nmfmt{\NAT@nm}%
   \NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi%
   \NAT@hyper@{\NAT@date}}
  {}% Do nothing if patch works
  {}% Do nothing if patch fails
\makeatother

@@

\newcommand{\Nc}{\langle N_c \rangle}
\newcommand{\Ns}{\langle N_s \rangle}
\newcommand{\Nt}{\langle N_t \rangle}
\newcommand{\halomod}{\textsc{halomod}}
\newcommand{\thm}{\textsc{TheHaloMod}}
\newcommand{\hmf}{\textsc{hmf}}
\newcommand{\framework}{\texttt{Framework}}
\newcommand{\component}{\texttt{Component}}
\newcommand{\parameter}{\texttt{parameter}}
\newcommand{\cached}{\texttt{cached\_quantity}}

\newcommand{\bd}[1]{\textcolor{purple}{\textbf{[BD: #1]}}}
\newcommand{\bde}[1]{\textcolor{blue}{#1}}
\newcommand{\sgm}[1]{\textcolor{green}{\textbf{[SM: #1]}}}
\newcommand{\ztc}[1]{\textcolor{cyan}{\textbf{[ZTC: #1]}}}

@@

%% Group authors per affiliation:
\author{Steven~G.~Murray$^a$, Benedikt Diemer$^b$, Zhaoting Chen$^c$}
\address{$^a$ Arizona State University, Tempe, AZ, USA \\
$^b$ Department of Astronomy, University of Maryland, College Park, MD 20742, USA\\
$^c$ Jodrell Bank Centre for Astrophysics, School of Physics and Astronomy, The University of Manchester, Manchester M13 9PL, UK}

\begin{abstract}
The halo model is a successful framework for describing the distribution of matter in the Universe -- from weak lensing observables to galaxy 2-point correlation functions. We review the basic formulation of the halo model and several of its components in the context of galaxy two-point statistics, developing a coherent framework for its application.

We use this framework to motivate the presentation of a new Python tool for simple and efficient calculation of halo model quantities, and their extension to galaxy statistics via a \textit{halo occupation distribution}, called \halomod. This tool is efficient, simple to use, comprehensive and importantly provides a great deal of flexibility in terms of custom extensions. 
This Python tool is complemented by a new web-application at \url{https://thehalomod.app} that supports the generation of many halo model quantities directly from the browser -- useful for educators, students, theorists and observers.

@@

The halo model \citep{Neyman1953,Peacock2000,Seljak2000,Ma2000,Cooray2002} is an enormously successful analytical description of the large-scale distribution of matter in our Universe.
%There are a diverse network of such models, but we focus here on a particularly successful description of the galaxy population, called the halo model \bd{the halo model does not describe galaxies! I think we should carefully distinguish these use cases. As you say in the next paragraph, the halo model must be combined with an HOD or other galaxy-halo model for actual predictions. That's why I'm actually not so sure about the first paragraph: it talks about modern survyes, but those are typically modeled with simulation + galaxy-halo model. So what is the real use case?} . 
It describes the statistics of the dark matter density field well into the nonlinear regime, beyond the reach of perturbation theory. 
It does so by combining linear theory predictions with empirical properties of dark matter halos, via the assumption that the sum total of dark matter resides in these clumps, and that a handful of simple functions based on the mass of these halos -- such as their radial density profile and clustering bias -- can universally describe them.

In combination with a \textit{halo occupation distribution} (HOD) model \citep{Kauffmann1997,Scoccimarro2001,Berlind2003,Zheng2005}, the predictions of the halo model can be extended to galaxy populations, and therefore used to model clustering in large galaxy surveys. 
One of the key advantages of halo model and HOD formalism is that they can predict any clustering statistic on any scale 
%is accessible through the halo model (or HOD) formalism 
\citep{Zehavi2011}, from real-space or projected two-point correlation functions (2PCFs), to galaxy-galaxy lensing, to higher-order correlations.

In practice, the HOD formalism has been widely used in the interpretation of galaxy populations in the past decade. Most of these studies have focused on determining the parameters of the HOD (i.e. the galaxy-halo connection) from the 2PCF of galaxies \citep[eg.][]{Moustakas2002,Bullock2002,Zheng2004,Zehavi2005,Blake2008,Zehavi2011,Beutler2013,Skibba2015} using the analytical framework that we present in this paper. However, other observables such as galaxy-galaxy lensing \citep{Mandelbaum2006}, the 2PCF of radio galaxies \citep{Wake2008a,Kim2011}, galaxy-quasar cross-correlations \citep{Shen2013}, near-UV cross-correlations \citep{Krause2012} and \textsc{Hi} intensity map cross-correlations \citep{Padmanabhan2016b,Wolz2019} have also received application through the same framework. Furthermore, several studies have successfully combined observables to break degeneracies \citep{Leauthaud2011,Leauthaud2012,More2013}, and the combination of the galaxy 2PCF with weak-lensing \citep{More2015}, stellar mass functions \citep{Coupon2015} and group mass-to-number ratios \citep{Reddick2014} has resulted in the ability to simultaneously constrain cosmological parameters along with those of the HOD.
%Several related approaches that employ different means for connecting galaxies to halos have been developed, including conditional luminosity functions \citep[CLF;][]{Yang2003,vandenBosch2003,Yang2005,Cooray2005,
%Cooray2005a,Cooray2006,vandenBosch2013,Cacciato2013} and subhalo abundance matching (SHAM; eg. \citet{Conroy2006}). \bd{This paragraph sort of comes out of nowhere, and feels a bit disconnected. Also, given the long refernece lists for HOD models, just citing one paper for SHAM might piss some people off... I'd suggest we take it out altogether. Then, the next paragraph properly connects to the previous one as well. Moreover, I don't think SHAM is ever applied to halo-model ``data''? I've only ever seen it applied to simulated catalogs.}

@@

%However, while the HM has become an increasingly popular tool, it is still difficult to find a publicly-available, stable and efficient code for its calculation. \bd{I think this will annoy anyone who has written such a code because those are very fundamental benchmarks. Maybe we could say that those codes are not fully flexible?} 
%To be clear, it is not that implementations do not exist 
There are a few publicly available implementations of the HM, many of which were used in the testing of our code although the development of most of them has been discontinued (e.g., \textsc{chomp}\footnote{\url{https://code.google.com/p/chomp/}  -- discontinued.}, \textsc{HMcode}\footnote{\url{https://github.com/alexander-mead/HMcode/}} and \textsc{AUM}\footnote{\url{https://github.com/surhudm/aum} -- discontinued.}).
%, nor that they are poor implementations in general (indeed, all of these were useful in testing our own code). 
However, it is our experience from that a remarkable number of 
%researchers are using the HM in various ways, but almost ubiquitously they find need to write their own code -- 
practitioners develop their own tools --- whether based heavily on existing (public or private) code or from the ground up.

@@

This paper presents a robust new implementation of the analytic HM, called \halomod,  that aims to fill this important gap, and be as generally useful as possible by adhering to the following principles:
\begin{itemize}
    \item \textbf{Intuitive.}  The API is well-specified and intuitive for the user, and exhaustively documented. We illustrate the simplicity of the usage of \halomod\ in \S\ref{sec:halomod:overview:usage} and note that full online documentation is available at \url{https://halomod.readthedocs.io}, including API specifications and examples/tutorials. In addition, installing \halomod\ is as simple as running \verb|pip install halomod|.

@@

Both philosophically and technically, \halomod\ inherits from the \hmf\ halo mass function package\footnote{\url{https://github.com/steven-murray/hmf}} which was first presented in \cite{Murray2013a}.
Many developments have occurred in \hmf\ since its first publication, and the technical framework of \halomod\ presented in this paper is essentially inherited from the updates in \hmf. 
Thus, this paper can also secondarily be considered an update of \hmf.

@@

In addition to \halomod\ and \hmf\ package, we also present a new web-application, \thm\footnote{\url{ https://thehalomod.app}}, which is able to generate full halo model quantities (eg. two-point correlation functions and galaxy power spectra) without ever having to install the Python package.
It is a successor to the popular \textsc{HMFcalc} \citep{Murray2013a} web-application, and includes the full range of functionality of \textsc{HMFcalc}.
The presentation of \thm\ completes our vision for \halomod\ with (v) a tool for educators to easily and interactively present cosmologically relevant quantities graphically.

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{thehalomod_app_schematic.pdf} 
    \caption{\textsc{TheHaloMod} in action. The main page (right panel) shows the currently computed models, with a selection of quantities to plot, and a number of actions available for each model (edit/clone/report bug/delete). The input form (two insets on the left) provides a wide range of options, including all of the components presented throughout this paper. Each component is defined in its own tab. Those shown here are cosmology and the HOD model.} \label{fig:webpage-image}
\end{figure*}

This paper is structured as follows: \S\ref{sec:DMHaloModel} details the theory of the HM particularly in the context of dark matter two-point statistics, collating the various components involved in a manner consistent with our implementation. Following this, we describe how to extend the halo model to tracer populations in \S\ref{sec:TracerHaloModel}. Then, we describe our code and its usage in  \S\ref{sec:halomod}, and in \S\ref{sec:applications} we present an illustrative example. In \S\ref{sec:future} we define a prospectus for the future, before summarising and concluding in \S\ref{sec:summary}.

Note that the code to produce all figures of different component models in this paper, as well as the example application, are available publicly as examples in \halomod's documentation\footnote{At \url{https://halomod.readthedocs.io/en/latest/examples/component-showcase.html} and \url{https://halomod.readthedocs.io/en/latest/examples/fitting.html} respectively.}. This paper refers to \textsc{hmf} version 3.2.1 and \textsc{halomod} version 2.0.0.

\section{The Dark Matter Halo Model}
\label{sec:DMHaloModel}
The broad assumption underpinning the HM is that in hierarchical structure formation scenarios, all mass is expected to be bound to halos at some scale\footnote{This assumption is clearly an approximation, since even cold dark matter (CDM) has a free-streaming scale in the early Universe below which we expect non-virialized mass \citep{Frenk2012,Schneider2014a}}. If this is the case, then the entire nonlinear density field may be reconstructed by summing contributions from the individual halos. If, in addition, we may describe the average radial density profile of halos as spherically-symmetric, with a shape that depends solely on the mass of the halo itself, we can write:
\begin{equation}
    \label{eq:hm_sum}
    \rho(\mathbf{x}) = \sum \rho_{h,i}(|\mathbf{x} - \mathbf{x_i}|,m_i)
\end{equation}
where $\rho_{h}(r|m)$ is the density of a halo with mass $m$ at radius $r$, and $x_i$ are the coordinates of the halo centres.

The application of the HM rests in converting the prem\-ise of Eq. \ref{eq:hm_sum} into a semi-analytic integration. This requires knowledge of four key components:
\begin{enumerate}
    \item The average radial density profile of halos, $\rho(r|m)$
    \item The abundance of halos of a given mass (termed the halo mass function, or HMF), $n(m)$
    \item The expected overdensity of halos of mass $m$, $\delta_h(\mathbf{x},m)$ in a region with a given overdensity of matter $\delta(\mathbf{x})$, called the \textit{halo bias}.
    \item A model for the linear spatial distribution of matter, $\delta(\mathbf{x})$.
\end{enumerate}
Furthermore, detailed calculations require the following extra components:
\begin{enumerate}
    \setcounter{enumi}{3}
    \item Additional parameters that describe the halo profile, most
    commonly parameterized as a concentration-mass relation, $c(m)$, linking the shape of a halo profile to its mass.
    \item A model for ``halo exclusion", which accounts for double-counting intra-halo correlations.
    \item To extend the calculations to galaxies, a distribution function for the occupation of halos by galaxies, as a function of halo mass, $N(m)$ (the HOD; cf. \S\ref{sec:TracerHaloModel}). 
\end{enumerate}

Although the halo model can be used to describe the density field at any level of the $n$-point hierarchy, its most common application to date has been to compute two-point statistics. Thus, in this outline, we shall focus on the framework as it pertains to the 2PCF.

Our aim in this section is to introduce the theory in a manner conducive to our implementation, so we shall cover each of the six components in turn following the presentation of the core framework.

Throughout the section, we discuss isotropic two-point statistics, which measure the over-density of pairs of points at a certain scalar separation. This can be formulated either in Euclidean space, as the correlation function $\xi(r)$, or in Fourier-space, as the power spectrum $P(k)$. 
The correlation function is defined as the excess probability of locating a particle at separation $\mathbf{r}$ from a particle at position $\mathbf{x}$ in a given spatial distribution, and is written (for a homogeneous universe)
\begin{equation}
    \xi(\mathbf{r}) = \frac{1}{V}\int_V d^3\mathbf{x}\ \delta(\mathbf{x}) \delta(\mathbf{x} - \mathbf{r}),
\end{equation}
where $\delta(\mathbf{x})$ is the overdensity
\begin{equation}
    \delta(\mathbf{x}) = \frac{\rho(\mathbf{x}) - \bar{\rho}}{\bar{\rho}}.
\end{equation}
The power spectrum is merely the Fourier transform of the correlation function, and the standard Fourier convention in cosmology renders it
\begin{equation}
    \xi(\mathbf{r}) = \int \frac{d^3 \mathbf{k}}{(2\pi)^3} P(\mathbf{k}) e^{i \mathbf{k}\cdot \mathbf{r}}.
\end{equation}
Under the assumption of cosmic isotropy, the correlation function and power spectrum depend only on the amplitude of $\mathbf{r}$ and $\mathbf{k}$, respectively. Transformation from an (isotropic) power spectrum $P(k)$ to an (isotropic) correlation function $\xi(r)$ can thus be performed using an order-$1/2$ \textit{Hankel} transform, which is the 3D Fourier transform of a spherically symmetric distribution:
\begin{equation}
\label{eq:hankel}
\xi(r) = \frac{1}{2\pi^2}\int_0^\infty P(k)k^2 j_0(kr)dk,
\end{equation}
where $j_0$ is the zeroth-order \textit{spherical} Bessel function,
\begin{equation}
    j_0(x) = \frac{\sin x}{x}.
\end{equation}
We note that the halo model is \textit{not limited} to such two-point statistics, but our implementation currently is.

\subsection{Clustering Framework}
\label{sec:haloclustering}
It is convenient to formulate the framework of clustering in two regimes \citep{Seljak2000}, intra- and inter-halo pairs, called the 1-halo and 2-halo terms respectively. These approximately correspond to small- and large-scale structure (i.e. the contribution of each in the opposite regime is negligible), where `small' is sub-megaparsec, and `large' is $>\sim5h^{-1}$Mpc.

@@

The total power spectrum and 2-point correlation function can be written simply as
\begin{equation}
    P(k) = P^{1h}(k) + P^{2h}(k),
\end{equation}
and
\begin{equation}
    \xi(r) = [1 + \xi^{1h}(r)] + \xi^{2h}(r).
\end{equation}
We will first describe the calculation of the 1-halo and 2-halo terms, each of which combines multiple physical components, which we will proceed to describe in 
\S\ref{sec:powerspec}-\ref{sec:theory:exclusion}.

>>>>

\usepackage{array}
\usepackage{lineno}
\usepackage{xcolor}
\usepackage{lineno}
\usepackage[colorlinks=true,urlcolor=black,linkcolor=blue,citecolor=blue]{hyperref}
\ifthenelse{\boolean{useminted}}{\usepackage{minted}}{}
\usepackage{longtable,booktabs,tabu}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage{pbox}
\usepackage{amssymb}
\usepackage{layouts}
\usepackage{subfig}
\usepackage{microtype}
\usepackage{listings}
\usepackage{xspace}
\usepackage[capitalise]{cleveref}
\usepackage[normalem]{ulem}

\usepackage{etoolbox}

\creflabelformat{equation}{#2\textup{#1}#3}

\makeatletter
% Patch case where name and year have no delimiter
\patchcmd{\NAT@citex}
  {\@citea\NAT@hyper@{\NAT@nmfmt{\NAT@nm}\NAT@date}}
  {\@citea\NAT@nmfmt{\NAT@nm}\NAT@hyper@{\NAT@date}}
  {}% Do nothing if patch works
  {}% Do nothing if patch fails
% Patch case where name and year have basic delimiter
\patchcmd{\NAT@citex}
  {\@citea\NAT@hyper@{%
     \NAT@nmfmt{\NAT@nm}%
     \hyper@natlinkbreak{\NAT@aysep\NAT@spacechar}{\@citeb\@extra@b@citeb}%
     \NAT@date}}
  {\@citea\NAT@nmfmt{\NAT@nm}%
   \NAT@aysep\NAT@spacechar%
   \NAT@hyper@{\NAT@date}}
  {}% Do nothing if patch works
  {}% Do nothing if patch fails
% Patch case where name and year are separated by a prenote
\patchcmd{\NAT@citex}
  {\@citea\NAT@hyper@{%
     \NAT@nmfmt{\NAT@nm}%
     \hyper@natlinkbreak{\NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi}%
       {\@citeb\@extra@b@citeb}%
     \NAT@date}}
  {\@citea\NAT@nmfmt{\NAT@nm}%
   \NAT@spacechar\NAT@@open\if*#1*\else#1\NAT@spacechar\fi%
   \NAT@hyper@{\NAT@date}}
  {}% Do nothing if patch works
  {}% Do nothing if patch fails
\makeatother

@@

\newcommand{\Nc}{\langle N_{\rm c} \rangle}
\newcommand{\Ns}{\langle N_{\rm s} \rangle}
\newcommand{\Nt}{\langle N_{\rm t} \rangle}
\newcommand{\halomod}{\textsc{halomod}\xspace}
\newcommand{\python}{\textsc{python}\xspace}
\newcommand{\thm}{\textsc{TheHaloMod}\xspace}
\newcommand{\hmf}{\textsc{hmf}\xspace}
\newcommand{\framework}{\texttt{Framework}\xspace}
\newcommand{\component}{\texttt{Component}\xspace}
\newcommand{\parameter}{\texttt{parameter}\xspace}
\newcommand{\cached}{\texttt{cached\_quantity}\xspace}

\newcommand{\bd}[1]{\textcolor{purple}{\textbf{[BD: #1]}}}
\newcommand{\bde}[1]{\textcolor{blue}{#1}}
\newcommand{\sgm}[1]{\textcolor{green}{\textbf{[SM: #1]}}}
\newcommand{\ztc}[1]{\textcolor{cyan}{\textbf{[ZTC: #1]}}}
\newcommand{\redbf}[1]{\textcolor{red}{\textbf{#1}}}
\newcommand{\red}[1]{\textcolor{red}{{#1}}}

@@

%% Group authors per affiliation:
\author{Steven~G.~Murray$^a$, Benedikt Diemer$^b$, Zhaoting Chen$^c$, Anton Glenn Neuhold Jr.$^d$, Alexander Schnapp$^d$, Tia Peruzzi$^d$, Daniel Blevins$^d$, Trent Engelman$^d$}
\address{$^a$ School of Earth and Space Exploration, Arizona State University, Tempe, AZ, USA \\
$^b$ Department of Astronomy, University of Maryland, College Park, MD 20742, USA\\
$^c$ Jodrell Bank Centre for Astrophysics, School of Physics and Astronomy, The University of Manchester, Manchester M13 9PL, UK \\
$^d$ ASU School of Computing, Informatics, and Decision Systems Engineering, Arizona State University, Tempe, AZ, USA \\
}

\begin{abstract}
The halo model is a successful framework for describing the distribution of matter in the Universe -- from weak lensing observables to galaxy n-point correlation functions. We review the basic formulation of the halo model and several of its components in the context of galaxy 2-point statistics, developing a coherent framework for its application.

We use this framework to motivate the presentation of a new \python tool for simple and efficient calculation of halo model quantities, and their extension to galaxy statistics via a \textit{halo occupation distribution}, called \halomod. This tool is efficient, simple to use, comprehensive and importantly provides a great deal of flexibility in terms of custom extensions. 
This \python tool is complemented by a new web-application at \url{https://thehalomod.app} that supports the generation of many halo model quantities directly from the browser -- useful for educators, students, theorists and observers.

@@

The halo model \citep{Neyman1953,Scherrer1991,Peacock2000,Seljak2000,Ma2000,Cooray2002} is an enormously successful analytical description of the large-scale distribution of matter in our Universe.
%There are a diverse network of such models, but we focus here on a particularly successful description of the galaxy population, called the halo model \bd{the halo model does not describe galaxies! I think we should carefully distinguish these use cases. As you say in the next paragraph, the halo model must be combined with an HOD or other galaxy-halo model for actual predictions. That's why I'm actually not so sure about the first paragraph: it talks about modern survyes, but those are typically modeled with simulation + galaxy-halo model. So what is the real use case?} . 
It describes the statistics of the dark matter density field well into the nonlinear regime, beyond the reach of perturbation theory. 
It does so by combining linear theory predictions with empirical properties of dark matter halos, via the assumption that the sum total of dark matter resides in these clumps, and that a handful of simple functions based on the mass of these halos -- such as their radial density profile and clustering bias -- can universally describe them.

In combination with a \textit{halo occupation distribution} (HOD) model \citep{Kauffmann1997,Scoccimarro2001,Berlind2003,Zheng2005}, the predictions of the halo model can be extended to galaxy populations, and therefore used to model clustering in large galaxy surveys. 
One of the key advantages of halo model and HOD formalism is that they can predict any clustering statistic on any scale 
%is accessible through the halo model (or HOD) formalism 
\citep{Zehavi2011}, from real-space or projected 2-point correlation functions (2PCFs), to galaxy-galaxy lensing, to higher-order correlations.

In practice, the HOD formalism has been widely used in the interpretation of galaxy populations in the past decade. Most of these studies have focused on determining the parameters of the HOD (i.e. the galaxy-halo connection) from the 2PCF of galaxies \citep[eg.][]{Moustakas2002,Bullock2002,Zheng2004,Zehavi2005,Blake2008,Zehavi2011,Beutler2013,Skibba2015,Rodriguez-Torres2015,Nicola2020,Zhou2021, Ishikawa2021} using the analytical framework that we present in this paper. However, other observables such as galaxy-galaxy lensing \citep{Mandelbaum2006,Cacciato2012,Clampitt2017,Dvornik2018}, the 2PCF of radio galaxies \citep{Wake2008a,Kim2011,Nusser2015}, galaxy-quasar cross-correlations \citep{Shen2013}, near-UV cross-correlations \citep{Krause2012}, Fast Radio Bursts \citep[FRBs;][]{Rafiei-Ravandi2020} and \textsc{Hi} intensity map cross-correlations \citep{Padmanabhan2016b,Wolz2019,Chen2021} and auto-correlations \citep{Schneider2020,Umeh2021} have also received application through the same framework. Furthermore, several studies have successfully combined observables (either in cross-correlation or via joint likelihoods) to break degeneracies \citep[eg.][]{Leauthaud2011,Leauthaud2012,More2013,Schaan2021}. 
Finally, the combination of the galaxy 2PCF with other statistics, eg. weak-lensing \citep{More2015,Miyatake2020}, cluster counts and auto-correlations \citep{To2021}, or both \citep{To2020}, stellar mass functions \citep{Coupon2015}
 and group mass-to-number ratios \citep{Reddick2014} has resulted in the ability to simultaneously constrain cosmological parameters along with those of the HOD.
%Several related approaches that employ different means for connecting galaxies to halos have been developed, including conditional luminosity functions \citep[CLF;][]{Yang2003,vandenBosch2003,Yang2005,Cooray2005,
%Cooray2005a,Cooray2006,vandenBosch2013,Cacciato2013} and subhalo abundance matching (SHAM; eg. \citet{Conroy2006}). \bd{This paragraph sort of comes out of nowhere, and feels a bit disconnected. Also, given the long refernece lists for HOD models, just citing one paper for SHAM might piss some people off... I'd suggest we take it out altogether. Then, the next paragraph properly connects to the previous one as well. Moreover, I don't think SHAM is ever applied to halo-model ``data''? I've only ever seen it applied to simulated catalogs.}

@@

%However, while the HM has become an increasingly popular tool, it is still difficult to find a publicly-available, stable and efficient code for its calculation. \bd{I think this will annoy anyone who has written such a code because those are very fundamental benchmarks. Maybe we could say that those codes are not fully flexible?} 
%To be clear, it is not that implementations do not exist 
There are a few publicly available implementations of the analytic HM, many of which were used in the testing of our code (e.g., \textsc{chomp}\footnote{\url{https://github.com/morriscb/chomp}  -- discontinued.}, \textsc{HMcode}\footnote{\url{https://github.com/alexander-mead/HMcode/}}, \textsc{CCL}\footnote{\url{https://github.com/LSSTDESC/CCL}} and \textsc{AUM}\footnote{\url{https://github.com/surhudm/aum}}).
In addition, methods of computing HM quantities based on emulation from simulations are becoming available (eg. \textsc{DarkQuest}\footnote{\url{https://github.com/DarkQuestCosmology/dark_emulator_public}} and \textsc{CosmicEmu}\footnote{\url{https://github.com/lanl/CosmicEmu}} -- though the latter only supports matter correlations).
%, nor that they are poor implementations in general (indeed, all of these were useful in testing our own code). 
However, it is our experience from that a remarkable number of 
%researchers are using the HM in various ways, but almost ubiquitously they find need to write their own code -- 
practitioners develop their own tools --- whether based heavily on existing (public or private) code or from the ground up. 
Additionally, existing codes -- where still in active development -- do not tend to offer a great deal of flexibility in the composition and extension of 
the various components and models involved in the halo model, rather focusing on a limited set of well-honed models and tuned for maximum performance.

@@

This paper presents a robust new implementation of the analytic HM, called \halomod\footnote{\url{https://github.com/halomod/halomod}},  that aims to fill this important gap, and be as generally useful as possible by adhering to the following principles:
\begin{itemize}
    \item \textbf{Intuitive.}  The API is well-specified and intuitive for the user, and exhaustively documented. We illustrate the simplicity of the usage of \halomod\ in \S\ref{sec:halomod:overview:usage} and note that full online documentation is available at \url{https://halomod.readthedocs.io}, including API specifications and examples/tutorials. In addition, installing \halomod\ is as simple as running \verb|pip install halomod|.

@@

Both philosophically and technically, \halomod\ inherits from the \hmf\ halo mass function package\footnote{\url{https://github.com/halomod/hmf}} which was first presented in \cite{Murray2013a}.
Many developments have occurred in \hmf\ since its first publication, and the technical framework of \halomod\ presented in this paper is essentially inherited from the updates in \hmf. 
Thus, this paper can also secondarily be considered an update of \hmf.

@@

In addition to \halomod\ and \hmf\ package, we also present a new web-application, \thm\footnote{\url{ https://thehalomod.app}}, which is able to generate full halo model quantities (eg. 2-point correlation functions and galaxy power spectra) without ever having to install the \python package.
It is a successor to the popular \textsc{HMFcalc} \citep{Murray2013a} web-application, and includes the full range of functionality of \textsc{HMFcalc}.
The presentation of \thm\ completes our vision for \halomod\ with (v) a tool for educators to easily and interactively present cosmologically relevant quantities graphically.

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth,clip=true,trim=0 2cm 0 1cm]{screenshot_thm.pdf} 
    \caption{\textsc{TheHaloMod} in action. The main page (right panel) shows the currently computed models, with a selection of quantities to plot, and a number of actions available for each model (edit/clone/report bug/delete). The plot itself is interactive. The input form (left) provides a wide range of options, including all of the components presented throughout this paper. Each component has its own section. The interface features responsive design and a dark mode.} \label{fig:webpage-image}
\end{figure*}

This paper is structured as follows: \S\ref{sec:DMHaloModel} details the theory of the HM particularly in the context of dark matter 2-point statistics, collating the various components involved in a manner consistent with our implementation. Following this, we describe how to extend the halo model to tracer populations in \S\ref{sec:TracerHaloModel}. Then, we describe our code and its usage in  \S\ref{sec:halomod}, and in \S\ref{sec:applications} we present an illustrative example. In \S\ref{sec:future} we define a prospectus for the future, before summarising and concluding in \S\ref{sec:summary}.

Note that the code to produce all figures of different component models in this paper, as well as the example application, are available publicly as examples in \halomod's documentation\footnote{At \url{https://halomod.readthedocs.io/en/latest/examples/component-showcase.html} and \url{https://halomod.readthedocs.io/en/latest/examples/fitting.html} respectively.}. This paper refers to \textsc{hmf} version 3.4.1 and \textsc{halomod} version 2.1.0.

\section{The Dark Matter Halo Model}
\label{sec:DMHaloModel}
The broad assumption underpinning the HM is that in hierarchical structure formation scenarios, all mass is expected to be bound to halos at some scale\footnote{This assumption is clearly an approximation, since even cold dark matter (CDM) has a free-streaming scale in the early Universe below which we expect non-virialized mass \citep{Frenk2012,Schneider2015}}. If this is the case, then the entire nonlinear density field may be reconstructed by summing contributions from the individual halos. If, in addition, we may describe the average radial density profile of halos as spherically-symmetric, with a shape that depends solely on the mass of the halo itself, we can write:
\begin{equation}
    \label{eq:hm_sum}
    \rho(\mathbf{x}) = \sum \rho_{h,i}(|\mathbf{x} - \mathbf{x_i}|,m_i)
\end{equation}
where $\rho_{h}(r|m)$ is the density of a halo with mass $m$ at radius $r$, and $x_i$ are the coordinates of the halo centres.

The application of the HM rests in converting the prem\-ise of \cref{eq:hm_sum} into a semi-analytic integration. This requires knowledge of four key components:
\begin{enumerate}
    \item The moments of the density profiles of halos, $\langle \rho(r|m) \rangle$, $\langle \rho^2_{h, i}(r|m) \rangle$, etc.
    \item The abundance of halos of a given mass (termed the halo mass function, or HMF), $n(m)$
    \item The expected overdensity of halos of mass $m$, $\Delta_{\rm h}(\mathbf{x},m)$ in a region with a given overdensity of matter $\delta(\mathbf{x})$, called the \textit{halo bias}.
    \item A model for the spatial distribution of matter, $\delta(\mathbf{x})$.
\end{enumerate}
Furthermore, detailed calculations require the following extra components:
\begin{enumerate}
    \setcounter{enumi}{4}
    \item Additional parameters that describe the halo profile, most
    commonly parameterized as a concentration-mass relation, $c(m)$, linking the shape of a halo profile to its mass.
    \item A model for ``halo exclusion", which accounts for double-counting intra-halo correlations.
    \item To extend the calculations to galaxies, a distribution function for the occupation of halos by galaxies, as a function of halo mass, $N(m)$ (the HOD; cf. \S\ref{sec:TracerHaloModel}). 
\end{enumerate}

Although the halo model can be used to describe the density field at any level of the $n$-point hierarchy, its most common application to date has been to compute 2-point statistics. Thus, in this outline, we shall focus on the framework as it pertains to the 2PCF.
In particular, the 2PCF only requires the first two moments of the density profile, and we will furthermore assume that the second moment is just the square of the average density profile (i.e. we only require the first moment, cf. \citealt{Giocoli2010}).
Our aim in this section is to introduce the theory in a manner conducive to our implementation, so we shall cover each of the six components in turn following the presentation of the core framework.

Throughout the section, we discuss isotropic 2-point statistics, which measure the over-density of pairs of points at a certain scalar separation. This can be formulated either in Euclidean space, as the correlation function $\xi(r)$, or in Fourier-space, as the power spectrum $P(k)$. 
The correlation function is defined as the excess probability of locating a particle at separation $\mathbf{r}$ from a particle at position $\mathbf{x}$ in a given spatial distribution, and is written (for a homogeneous universe)
\begin{equation}
    \xi(\mathbf{r}) = \frac{1}{V}\int_V {\rm d}^3\mathbf{x}\ \delta(\mathbf{x}) \delta(\mathbf{x} - \mathbf{r}),
\end{equation}
where $\delta(\mathbf{x})$ is the overdensity
\begin{equation}
    \delta(\mathbf{x}) = \frac{\rho(\mathbf{x}) - \bar{\rho}}{\bar{\rho}}.
\end{equation}
The power spectrum is merely the Fourier transform of the correlation function, and the standard Fourier convention in cosmology renders it
\begin{equation}
    \xi(\mathbf{r}) = \int \frac{{\rm d}^3 \mathbf{k}}{(2\pi)^3} P(\mathbf{k}) e^{i \mathbf{k}\cdot \mathbf{r}}.
\end{equation}
Under the assumption of cosmic isotropy, the correlation function and power spectrum depend only on the amplitude of $\mathbf{r}$ and $\mathbf{k}$, respectively. Transformation from an (isotropic) power spectrum $P(k)$ to an (isotropic) correlation function $\xi(r)$ can thus be performed using an order-$1/2$ \textit{Hankel} transform, which is the 3D Fourier transform of a spherically symmetric distribution:
\begin{equation}
\label{eq:hankel}
\xi(r) = \frac{1}{2\pi^2}\int_0^\infty P(k)k^2 j_0(kr){\rm d}k,
\end{equation}
where $j_0$ is the zeroth-order \textit{spherical} Bessel function,
\begin{equation}
    j_0(x) = \frac{\sin x}{x}.
\end{equation}
We note that the halo model is \textit{not limited} to such 2-point statistics, but our implementation currently is.

\subsection{Clustering Framework}
\label{sec:haloclustering}
It is convenient to formulate the framework of clustering in two regimes \citep{Seljak2000}, intra- and inter-halo pairs, called the 1-halo and 2-halo terms respectively. These approximately correspond to small- and large-scale structure (i.e. the contribution of each in the opposite regime is negligible), where `small' is sub-megaparsec, and `large' is $\gtrsim5h^{-1}$Mpc.

@@

The total power spectrum and 2-point correlation function can be written simply as
\begin{equation}
    P(k) = P^{\rm 1h}(k) + P^{\rm 2h}(k),
\end{equation}
and
\begin{equation}
    \xi(r) = [1 + \xi^{\rm 1h}(r)] + \xi^{\rm 2h}(r).
\end{equation}
We will first describe the calculation of the 1-halo and 2-halo terms, each of which combines multiple physical components, which we will proceed to describe in 
\S\ref{sec:powerspec}-\ref{sec:theory:exclusion}.
